# Session 2.2 {#session-2-2}

```{r echo=FALSE}
library(patchwork)
```


## Test your understanding 1

```{r, include = FALSE}
bar <- c(answer = "geom_bar", x = "geom_col")
col <- c(x = "geom_bar", answer = "geom_col")
```

Here is a small data table.

country             | population      | island
:-------------------|----------------:|:--------------
Northern Ireland    | 1,895,510       | Ireland
Wales               | 3,169,586       | Great Britain
Republic of Ireland | 4,937,786       | Ireland
Scotland            | 5,466,000       | Great Britain
England             | 56,550,138      | Great Britain

::: {.try data-latex=""}

* What geom would you use to plot the population for each of the 5 countries? `r mcq(col)`
```{r test-counting-categories, echo = FALSE, results='asis'}
opt <- c(answer = "aes(x = country, y = population)",
         x = "aes(x = population, y = country)",
         x = "aes(x = country)",
         x = "aes(x = island)",
         x = "aes(y = population)")
cat("* What mapping would you use? ", longmcq(opt))
```

* What geom would you use to plot the number of countries on each island? `r mcq(bar)`
```{r, echo = FALSE, results='asis'}
opt <- c(x = "aes(x = country, y = population)",
         x = "aes(x = population, y = country)",
         x = "aes(x = country)",
         answer = "aes(x = island)",
         x = "aes(y = population)")
cat("* What mapping would you use? ", longmcq(opt))
```
:::

## Test your understanding 2

Imagine you have a table of the [population for each country in the world](https://population.un.org/wpp/Download/Standard/Population/){target="_blank"} with the columns `country` and `population`. We'll just look at the 76 countries with populations of less than a million.

```{r test-one-continuous, echo = FALSE}
# set up quiz answers
hist <- c(answer = "geom_histogram", x = "geom_freqpoly", x = "geom_density")
freq <- c(x = "geom_histogram", answer = "geom_freqpoly", x = "geom_density")
dens <- c(x = "geom_histogram", x = "geom_freqpoly", answer = "geom_density")

# load data
pop_data <- readxl::read_excel("data/WPP2019_POP_F01_1_TOTAL_POPULATION_BOTH_SEXES.xlsx", skip = 16) %>%
  filter(Type == "Country/Area") %>%
  select(country = 3, population = `2020`) %>%
  mutate(population = round(as.numeric(population) * 1000)) %>%
  filter(population < 1e6)
  
# make plots
g <- ggplot(pop_data, aes(x = population)) +
  scale_x_continuous(breaks = seq(0, 1e6, 1e5),
                     labels = c(paste0(0:9*100, "K"), "1M")) +
  scale_y_continuous(name = "")

hist_plot <- g + geom_histogram(binwidth = 1e5, boundary = 0, fill = "white", color = "black")
freq_plot <- g + geom_freqpoly(binwidth = 1e5, boundary = 0, color = "black") +
  coord_cartesian(xlim = c(0, 1e6))
dens_plot <- g + geom_density(color = "black", fill = "white")

dens_plot + hist_plot + freq_plot + 
  plot_layout(nrow = 3) +
  plot_annotation(tag_levels = "A")

```

::: {.try data-latex=""}

* What kind of plot is A? `r mcq(dens)`
* What kind of plot is B? `r mcq(hist)`
* What kind of plot is C? `r mcq(freq)`
```{r, echo = FALSE, results='asis'}
opts <- c(x = "aes(x = country, y = population)",
          x = "aes(x = population, y = country)",
          answer = "aes(x = population)",
          x = "aes(x = population, y = count)")
cat("* How would you set the mapping for these plots? ", longmcq(opts))
```
* What is the `binwidth` of the histogram? `r mcq(c("1", "100", answer = "100K", "1M"))`
:::

## Test your understanding 3

```{r test-grouped-continuous, echo=FALSE, fig.height = 2.5, message=FALSE, warning=FALSE}
survey_data <- read_csv("https://psyteachr.github.io/ads-v1/data/survey_data.csv")
box <- c(x = "geom_box()",
         answer = "geom_boxplot()",
         x = "geom_violin()",
         x = "geom_violinplot()")

violin <- c(x = "geom_box()",
         x = "geom_boxplot()",
         answer = "geom_violin()",
         x = "geom_violinplot()")

median <- c(x = "mean",
            answer = "median",
            x = "mode")

e04 <- setNames(faux::make_id(10, "e", 2), rep("x", 10))
e07 <- e04
names(e04)[4] <- "answer"
names(e07)[7] <- "answer"

aes <- sample(c(
  answer = "aes(x = employee_id, y = call_time, fill = employee_id)",
  x = "aes(x = employee_id, y = call_time, fill = call_time)",
  x = "aes(x = employee_id, y = call_time, colour = call_time)",
  x = "aes(x = employee_id, y = call_time, colour = employee_id)"
))

g <- ggplot(survey_data, aes(x = employee_id, y = call_time, fill = employee_id)) +
  guides(fill = "none")
  
v <- g + geom_violin(draw_quantiles = 0.5)
b <- g + geom_boxplot()

v + b + plot_annotation(tag_levels = "A")
```

::: {.try data-latex=""}
* How would you create plot A? `r mcq(violin)`
* How would you create plot B? `r mcq(box)`
* What does the mapping look like for both plots? `r longmcq(aes)`
* Which employee tends to have the longest calls? `r mcq(e04)`
* Which employee has the record longest call? `r mcq(e07)`
:::

## Exercises

To consolidate what you've learned so far, use the built-in `starwars` data set to create each of the following:

* A histogram of heights
* A scatterplot of height v mass with line of best fit
* A bar chart of counts for gender

`r hide()`

```{r}
# load data
data("starwars")

# histogram
ggplot(starwars, aes(x = height)) +
  geom_histogram()

# scatterplot
ggplot(starwars, aes(x = height, y = mass)) +
  geom_point() +
  geom_smooth()

# bar chart
ggplot(starwars, aes(x = gender)) +
  geom_bar()

```

`r unhide()`

Then use the built-in`gss_cat` data set to create:

* A grouped density plot of age by marital status
* A frequency plot of age by marital status
* A violin-boxplot of age by marital status


`r hide()`

```{r}

# load data
data("gss_cat")

# grouped density
ggplot(gss_cat, aes(x = age, fill = marital)) +
  geom_density(alpha = .6)

# frequency plot
ggplot(gss_cat, aes(x = age, colour = marital)) +
  geom_freqpoly()

# violin-boxplot
ggplot(gss_cat, aes(x = marital, y = age)) +
  geom_violin() +
  geom_boxplot(width = .2)

```

`r unhide()`

Finally, pick your two favourite plots from the above, spend some time adjusting the aesthetics (colours, themes, labels etc.), then combine them using `r pkg("patchwork")`. If you're on Twitter, share them with #PsyTeachR and tag [emilynordmann](https://twitter.com/emilynordmann).