# Session 2.2 {#session-2-2}

## Exercises

To consolidate what you've learned so far, use the built-in `starwars` data set to create each of the following:

* A histogram of heights
* A scatterplot of height v mass with line of best fit
* A bar chart of counts for gender

`r hide()`

```{r}
# load data
data("starwars")

# histogram
ggplot(starwars, aes(x = height)) +
  geom_histogram()

# scatterplot
ggplot(starwars, aes(x = height, y = mass)) +
  geom_point() +
  geom_smooth()

# bar chart
ggplot(starwars, aes(x = gender)) +
  geom_bar()

```

`r unhide()`

Then use the built-in`gss_cat` data set to create:

* A grouped density plot of age by marital status
* A frequency plot of age by marital status
* A violin-boxplot of age by marital status


`r hide()`

```{r}

# load data
data("gss_cat")

# grouped density
ggplot(gss_cat, aes(x = age, fill = marital)) +
  geom_density(alpha = .6)

# frequency plot
ggplot(gss_cat, aes(x = age, colour = marital)) +
  geom_freqpoly()

# violin-boxplot
ggplot(gss_cat, aes(x = marital, y = age)) +
  geom_violin() +
  geom_boxplot(width = .2)

```

`r unhide()`

Finally, pick your two favourite plots from the above, spend some time adjusting the aesthetics (colours, themes, labels etc.), then combine them using `r pkg("patchwork")`. If you're on Twitter, share them with #PsyTeachR and tag [emilynordmann](https://twitter.com/emilynordmann).